---
title: "Transforming Wide Datasets"
author: "Jeff Parks"
date: "`r Sys.Date()`"
description: "CUNY Data 607 - Fall 2021, Project 2"
output:
#  pdf_document:
#    latex_engine: xelatex
#    extra_dependencies: ["geometry", "multicol", "multirow"]
#  prettydoc::html_pretty:
#    theme: architect
#    highlight: github
  rmdformats::robobook:
    highlight: kate
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
library(kableExtra)
library(readxl)
```

## Assignment

Three different untidy (or "wide") datasets were selected in order to practice different methods of data preparation.

## Dataset 1: World Governmental Indicators

The source 

### Data Prep
```{r, message=FALSE, warning=FALSE}
# import excel sheet and skip header rows
df_raw <- read_excel('https://github.com/jefedigital/cuny-data-607-projects/blob/main/wide-datasets/data/wgi/wgidataset.xlsx?raw=true', 
                     sheet='ControlofCorruption', skip=12)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
kable(df_raw[1:5,1:4], caption="raw df") %>%
  kable_minimal()
```

discussion

```{r, message=FALSE, warning=FALSE}
# convert the column names and first row into 2 vectors
df_col_1 <- colnames(df_raw)
df_col_2 <- as.character(slice(df_raw,1))

# get rid of the auto numbering in the column names
df_col_1 <- str_remove(df_col_1, '\\..*$')

# now join each element of the two vectors with paste0 (no separators)
df_cols <- paste0(df_col_1, df_col_2)

# make new df with our new column names and the values
df <- df_raw[-1,]
names(df) <- df_cols

# data cleanup .. replace all the instances of string '#N/A' with NA
nona <- function(s) {str_replace(s,'#N/A','NA')}
df <- df %>% mutate(across(everything(), nona))

# convert columns to numeric type
df <- type_convert(df)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
kable(df[1:5,1:5], caption="formatted df") %>%
  kable_minimal()
```

Pivot time .. we want seven cols in total: country, code, year, category and measure.

```{r, message=FALSE, warning=FALSE}
# we can use the names_pattern parameter of pivot_longer to separate Year and Category into separate columns .. pretty useful!
df <- df %>% pivot_longer(
 cols = `1996Estimate`:`2020Upper`,
 names_to = c('Year','Category'),
 names_pattern = '(.{4})(.*)',
 values_to = 'Measure'
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
kable(df[50:55,], caption="pivoted df") %>%
  kable_minimal()
```

### Analysis

Cool, now let's demonstrate a couple of simple visualizations...

```{r, echo=FALSE, fig.show='hold', out.width='50%'}

df_rank_2020 <- df %>% 
  filter(Category=='Rank', Year==2020) %>% 
  filter(Measure >= 90) %>%
  arrange(desc(Measure))

df_rank_2020 %>% ggplot(aes(x=reorder(Code, -Measure), y=Measure)) +
  geom_bar(stat='identity')
```

```{r, echo=FALSE, fig.show='hold', out.width='50%'}

df_rank_annual <- df %>% 
  filter(Category=='Rank', Code=='AFG') %>% 
  arrange(Year)

df_rank_annual %>% ggplot(aes(x=Year, y=Measure)) +
  geom_bar(stat='identity')

```

---

## Dataset 2: League of Legends Champions

### Data Prep
```{r, message=FALSE, warning=FALSE}
library(tidyjson)
library(rvest)

# get json from webpage with rvest read_html
page <- read_html('https://ddragon.leagueoflegends.com/cdn/11.19.1/data/en_US/champion.json')

# parse the html with rvest, return just the contents of the <body> element
json <- page %>% html_elements("body") %>% html_text()

# check out the structure
json %>% gather_object %>% json_types

# get only the 'data' object from the json and spread_all into a 
# json table, then convert back to regular dataframe
champions <- json %>% 
  enter_object(data) %>% 
  gather_object %>% 
  spread_all %>% 
  as_data_frame.tbl_json %>%
  select(!blurb) # long strings
```

```{r, message=FALSE, warning=FALSE}
kable(head(champions), caption="formatted df") %>%
  kable_minimal()
```

```{r, echo=FALSE, fig.show='hold', out.width='50%'}
df <- champions %>% 
  select(c('name','stats.hp','stats.mp','stats.movespeed','stats.armor'))

ggplot(df) +
 aes(x = stats.hp) +
 geom_histogram(binwidth = 10, color = "#112446", fill="#ffffff") +
 theme_minimal()

ggplot(df) +
  aes(x = stats.mp) +
  geom_histogram(binwidth=50, color = "#112446", fill="#ffffff") +
  theme_minimal()
```

---

## Dataset 3: Global Infant Mortality Rates

### Data Prep
```{r, message=FALSE, warning=FALSE}
df_mort <- read_csv('data/worldbank/API_SH.DYN.MORT_DS2_en_csv_v2_3012069.csv', skip=4)
df_meta <- read_csv('data/worldbank/Metadata_Country_API_SH.DYN.MORT_DS2_en_csv_v2_3012069.csv')

# cleanup
df_mort <- df_mort %>% 
  select(!('2020':'...66')) %>% 
  select(!('Indicator Name':'Indicator Code')) %>% 
  rename(Name = 'Country Name', Code = 'Country Code')

df_meta <- df_meta %>%
  select(!SpecialNotes) %>%
  select(!last_col())

# impute missing Region values
df_meta <- df_meta %>% 
  mutate(`Record Type` = ifelse(is.na(Region),'Region','Country')) %>%
  mutate(Region = ifelse(is.na(Region), TableName, Region)) %>% 
  rename(Code = 'Country Code') %>%
  select(!TableName)
```

```{r, message=FALSE, warning=FALSE}
# pivot
df_mort <- df_mort %>% 
  pivot_longer(!c('Name','Code'), 
               names_to = 'Year', 
               values_to = 'Mortality Rate')

# join
df <- left_join(df_mort, df_meta, by='Code')

# correct to actual rate (/1000)
df <- df %>% mutate(`Mortality Rate` = `Mortality Rate`/1000)

# analysis
df_regions <- df %>% filter(`Record Type` == 'Region')
df_countries <- df %>% filter(`Record Type` == 'Country')

unique(df_regions$Name)
unique(df_countries$Region)

```

```{r, message=FALSE, warning=FALSE}
df_countries_latam <- df_countries %>%
  filter(Region == 'Latin America & Caribbean')


ggplot(df_countries_latam) +
  aes(x = Year, y = `Mortality Rate`, group = Name, label=Name) +
  geom_line() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90))
```